# Gsynth {#sec-gsynth}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 7)
```

This chapter demonstrates the default **gsynth** estimator (`estimator = "gsynth"`, [@Xu2017]), which estimates latent factors from the control group only and then imputes counterfactuals for treated units. We cover three settings: block DID, staggered DID, and unbalanced panels.

```{r}
library(gsynth)
library(panelView)
data(gsynth)
```

## Block DID

We start with a simulated dataset: 5 treated units, 45 control units, and 30 time periods. Treatment begins at Period 21 for all treated units.

```{r}
head(simdata)
```

### Visualizing data

Before estimation, visualize the data structure with **panelView**.

```{r simdata-panelview-raw, fig.height = 10}
panelview(Y ~ D, data = simdata, index = c("id", "time"), 
          pre.post = TRUE)
```

Outcome trajectories by treatment status:

```{r simdata-panelview-outcome}
panelview(Y ~ D, data = simdata, index = c("id", "time"), 
          type = "outcome")
```

### Estimation & Inference

We estimate the model using the outcome $Y$, the treatment indicator $D$, and two covariates $X_1$ and $X_2$.

The `force` option specifies fixed effects: `"none"`, `"unit"`, `"time"`, or `"two-way"`. Cross-validation (`CV = TRUE`) selects the number of factors within the range specified by `c(0, 5)`.

For finite sample inference, set `se = TRUE`. With few treated units, use `inference = "parametric"`. 
```{r sim1, cache = TRUE}
system.time(
    out <- gsynth(Y ~ D + X1 + X2, data = simdata,
                  index = c("id", "time"), force = "two-way",
                  CV = TRUE, r = c(0, 5), se = TRUE,
                  inference = "parametric", nboots = 1000,
                  parallel = FALSE)
)
```

For super-population inferences (with a large $N$, and $N_{tr}$ grows with $N$), users can also use `inference = "nonparametric"` or `inference = "jackknife"`.

The output shows: `sigma2` (error variance), `IC` (information criterion), and `MSPE` (mean squared prediction error). Cross-validation selects the $r$ that minimizes MSPE.

```{r}
out$est.att
out$est.avg
out$est.beta
```

### Parallel computing

Parallel computing (`parallel = TRUE`) speeds up the bootstrap procedure. We recommend that users manually set the number of cores using the `cores` option. If this is not supplied or is `NULL`, we will automatically select the smaller of `8` and the number of usable system cores minus `2`, to prevent excessive use of system resources.

```{r sim2, cache = TRUE}
system.time(
out <- gsynth(Y ~ D + X1 + X2, data = simdata,
              index = c("id", "time"), force = "two-way",
              CV = TRUE, r = c(0, 5), se = TRUE,
              inference = "parametric", nboots = 1000,
              parallel = TRUE, cores = 10)
)
print(out)
```


### Cumulative effects

The `effect` function calculates cumulative treatment effects. Specify a period of interest with `period`; if omitted, all post-treatment periods are used.

```{r cumu1, cache = TRUE}
cumu1 <- effect(out, cumu = TRUE, id = NULL)
cumu1$effect.est.att
plot(cumu1)
```

Sub-group effects and average (non-cumulative) effects:

```{r cumu2, cache = TRUE}
cumu2 <- effect(out, cumu = FALSE, id = c(101, 102, 103), 
                period = c(1, 5))
cumu2$effect.est.att
plot(cumu2)
```

### Plot results

The default `plot` produces a "gap" plot showing the estimated ATT by period. The true effects in `simdata` go from 1 to 10 (plus noise) from period 21 to 30.

```{r sim-gap1}
plot(out)
```

A default *ggplot2* theme:

```{r sim-gap1a}
plot(out, theme.bw = FALSE)
```

Customize axes and title:

```{r sim-gap2}
plot(out, type = "gap", ylim = c(-10, 12), xlab = "Period",
     main = "My GSynth Plot")
```

#### Raw data

```{r sim-raw1}
plot(out, type = "raw")
```

#### Counterfactuals

```{r ct1}
plot(out, type = "counterfactual", raw = "none", main = "")
```

Without post-treatment shading:

```{r ct1a}
plot(out, type = "ct", raw = "none", main = "",
     shade.post = FALSE)
```

With quantile bands (5th--95th percentile):

```{r ct2}
plot(out, type = "counterfactual", raw = "band",
     xlab = "Time", ylim = c(-5, 35))
```

With all raw data:

```{r ct3}
plot(out, type = "counterfactual", raw = "all")
```

Unit-level counterfactuals:

```{r ct4}
plot(out, type = "counterfactual", id = 102)
```

```{r ct5}
plot(out, type = "counterfactual", id = 104,
     raw = "band", ylim = c(-10, 30))
```

```{r ct6}
plot(out, type = "counterfactual", id = 105,
     raw = "all", legendOff = TRUE)
```

#### Factors and loadings

```{r sim-factors, results='hide'}
plot(out, type = "factors", xlab = "Time")
```

```{r sim-loadings, fig.height = 5, results='hide'}
plot(out, type = "loadings")
```

## Staggered DID

The `turnout` dataset examines the effect of Election Day Registration (EDR) reforms on voter turnout in the United States, where treatment kicks in at different times.

```{r}
names(turnout)
```

### Data structure

The balanced panel has 9 treated units with staggered treatment timing.

```{r turnout-obs, fig.height = 10}
panelview(turnout ~ policy_edr, data = turnout,
          index = c("abb", "year"), pre.post = TRUE,
          by.timing = TRUE)
```

Outcome trajectories by treatment-status group:

```{r turnout-obs2, fig.height = 10}
panelview(turnout ~ policy_edr, data = turnout,
          index = c("abb", "year"), type = "outcome",
          main = "EDR Reform and Turnout",
          by.group = TRUE)
```

### Estimation without factors

When no factor is assumed, the estimates are close to difference-in-differences:

```{r turnout-did, cache = TRUE}
out0 <- gsynth(turnout ~ policy_edr + policy_mail_in + policy_motor,
               data = turnout, index = c("abb", "year"),
               se = TRUE, inference = "parametric",
               r = 0, CV = FALSE, force = "two-way",
               nboots = 1000, seed = 02139)
```

```{r turnout-did-gap}
plot(out0, type = "gap", xlim = c(-10, 5), ylim = c(-30, 20))
```

### Estimation with factors

Cross-validation selects the number of factors:

```{r turnout-est, cache = TRUE}
out <- gsynth(turnout ~ policy_edr + policy_mail_in + policy_motor,
              data = turnout, index = c("abb", "year"),
              se = TRUE, inference = "parametric",
              r = c(0, 5), CV = TRUE, force = "two-way",
              nboots = 1000, seed = 02139)
```

### Implied weights

`out$wgt.implied` ($N_{co} \times N_{tr}$) stores the implied weights of all control units for each treated unit. Unlike synthetic control, the weights can be both positive and negative.

```{r}
dim(out$wgt.implied)
sort(out$wgt.implied[, 8])
```

### Plot results

Gap plot:

```{r turnout-gap}
plot(out, type = "gap", xlim = c(-10, 5), ylim = c(-15, 15))
```

State-specific gap (Wisconsin):

```{r turnout-gap2}
plot(out, type = "gap", id = "WI", main = "Wisconsin")
```

Raw turnout data:

```{r turnout-raw}
plot(out, type = "raw", xlab = "Year", ylab = "Turnout")
```

Estimated average counterfactual (realigned by treatment timing):

```{r turnout-ct}
plot(out, type = "counterfactual", raw = "all")
```

Unit-level counterfactuals:

```{r turnout-ct1}
plot(out, type = "counterfactual", id = "CT")
```

```{r turnout-ct2}
plot(out, type = "counterfactual", id = "WY",
     raw = "all", legendOff = TRUE)
```

```{r turnout-ct3}
plot(out, type = "counterfactual", id = "WI",
     raw = "none", shade.post = FALSE, ylim = c(0, 100),
     legend.labs = c("Wisconsin Actual", "Wisconsin Counterfactual"))
```

Estimated factors and loadings:

```{r turnout-factors, results='hide'}
plot(out, type = "factors", xlab = "Year")
```

```{r turnout-loadings, fig.height = 5, results='hide'}
plot(out, type = "loadings")
```

## Unbalanced panels

**gsynth** can accommodate unbalanced panels. We demonstrate by randomly removing observations from the `turnout` dataset.

```{r}
set.seed(123456)
turnout.ub <- turnout[-c(which(turnout$abb == "WY")[1:15],
                         sample(1:nrow(turnout), 50, 
                                replace = FALSE)), ]
```

Visualize the data structure. White cells represent missing values.

```{r turnout-ub-panelview, fig.height = 10}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout.ub, index = c("abb", "year"),
          pre.post = TRUE)
```

### Estimation

```{r turnout-ub-est, cache = TRUE}
out <- gsynth(turnout ~ policy_edr + policy_mail_in + policy_motor,
              data = turnout.ub, index = c("abb", "year"),
              se = TRUE, inference = "parametric",
              r = c(0, 5), CV = TRUE, force = "two-way",
              parallel = TRUE, min.T0 = 8,
              nboots = 1000, seed = 02139)
```

### Plot results

The `"missing"` plot shows the data used in estimation. Gray cells with red dots indicate observations removed due to insufficient pre-treatment periods. The matrix is available in `out$obs.missing`, where 0 = missing, 1 = control, 2 = treated (pre), 3 = treated (post), 4 = treated (removed).

```{r turnout-ub-missing, fig.height = 10}
plot(out, type = "missing")
```

Subset to treated units only:

```{r turnout-ub-missing2}
plot(out, type = "missing", xlab = "Year", ylab = "State",
     main = "Treatment Status", id = out$id.tr,
     xlim = c(1920, 2012), axis.adjust = TRUE)
```

Gap plot with the unbalanced panel:

```{r turnout-ub-gap}
plot(out, type = "gap", xlim = c(-10, 5), ylim = c(-15, 15))
```

### Notes on performance

1. Unbalanced panels take significantly more time than balanced panels (often 100:1 or higher) due to the EM algorithm for missing values.

2. Adding covariates also slows the algorithm because the IFE/MC model takes more iterations to converge.

3. Setting `min.T0` to a positive number helps --- the algorithm discards treated units with too few pre-treatment periods. A larger $T_0$ reduces bias and extrapolation risk. For cross-validation, `min.T0` must be at least `r.max + 2`.
