# IFE-EM & MC {#sec-ife-mc}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 7)
```

::: {.callout-warning}

## Soft-deprecated in v1.5.0

`estimator = "ife"` (IFE-EM, @Gobillon2016), `estimator = "mc"` (matrix completion, @Athey2021), and the legacy `EM = TRUE` flag are soft-deprecated in `gsynth` v1.5.0 and will be **removed in v2.0.0** (target: late 2026).

Each call to one of these paths emits a one-time-per-session console message. Behavior is unchanged from v1.4.0 --- the deprecation is a signal, not a break.

For new code, use [`fect::fect()`](https://yiqingxu.org/packages/fect/) directly. The **gsynth** package is built around the GSC estimator (@Xu2017); IFE-EM and matrix completion are now documented and maintained in **fect**. Migration recipes are in @sec-migration below.

:::

In the difference-in-differences setting (with a large $N$, and $N_{tr}$ growing with $N$), @Gobillon2016 propose an EM algorithm that exploits pre-treatment observations from treated units; @Athey2021 propose a matrix-completion approach that imputes counterfactual outcomes under a low-rank structure. Both methods use information from never-treated controls and not-yet-treated observations of treated units, and conduct inference under a super-population framework.

For a full treatment, see the **fect** [User Manual](https://yiqingxu.org/packages/fect/).

This chapter documents the v1.5.0 wrapper paths for users with existing code that depends on `estimator = "ife"` or `estimator = "mc"`, and shows how to run a placebo test on the resulting fit. The chapter ends with migration recipes that move existing calls to `fect::fect()` directly.

## A working IFE-EM fit

```{r}
library(gsynth)
data(gsynth)
```

```{r sim-ife, cache = TRUE}
out_ife <- gsynth(Y ~ D + X1 + X2, data = simdata,
                  index = c("id", "time"), estimator = "ife",
                  force = "two-way", inference = "bootstrap",
                  se = TRUE, nboots = 500, r = c(0, 5),
                  CV = TRUE, parallel = TRUE, cores = 16)
print(out_ife)
```

```{r sim-ife-plot}
plot(out_ife, main = "Estimated ATT (IFE-EM)")
```

## A working matrix-completion fit

The `lambda` argument controls the regularization parameter. If left unset, a sequence of candidates is generated and selected via cross-validation. Set the candidate-list length with `nlambda`, or supply your own sequence via `lambda`. The number of CV folds is set with `k`.

```{r sim-mc, cache = TRUE}
out_mc <- gsynth(Y ~ D + X1 + X2, data = simdata,
                 estimator = "mc", index = c("id", "time"),
                 se = TRUE, nboots = 500, r = c(0, 5),
                 CV = TRUE, force = "two-way",
                 parallel = TRUE, cores = 16,
                 inference = "bootstrap")
```

```{r sim-mc-plot}
plot(out_mc, main = "Estimated ATT (MC)")
```

::: {.callout-important}

## Hard gate on parametric × IFE-EM (breaking change in v1.5.0)

In v1.4.0, `inference = "parametric"` combined with `estimator = "ife"` or `"mc"` was silently coerced to bootstrap inference with a warning. **In v1.5.0 this combination errors instead.**

```{r hard-gate, eval = FALSE}
## v1.4.0 (worked, with a warning):
gsynth(Y ~ D, data = ..., estimator = "ife",
       inference = "parametric", se = TRUE, nboots = 200)

## v1.5.0 (errors). Migrate to:
gsynth(Y ~ D, data = ..., estimator = "ife",
       inference = "bootstrap", se = TRUE, nboots = 200)
```

The hard gate matches **fect** v2.2.0+. The reason is that silent coercion can hide an inferential issue from users who deliberately chose IFE-EM: the parametric bootstrap is designed for the small-$N_{tr}$ GSC setting and is not a sensible choice once you switch to a super-population estimator. A hard error forces a conscious choice. For the longer discussion, see the **fect** [User Manual](https://yiqingxu.org/packages/fect/).

:::

## Placebo tests for IFE-EM and MC {#sec-ife-mc-placebo}

The basic placebo workflow is in [Chapter @sec-placebo]. We repeat it here on an IFE-EM fit because the placebo test is especially worth running on these estimators: the EM step blends pre-treatment treated cells into the factor estimates, and the matrix-completion regularizer trades bias for variance, so both can produce non-zero pre-treatment "effects" without a real signal.

```{r placebo-ife, cache = TRUE}
out_ife_pb <- gsynth(Y ~ D + X1 + X2, data = simdata,
                     index = c("id", "time"), estimator = "ife",
                     force = "two-way", CV = FALSE, r = 2,
                     se = TRUE, inference = "bootstrap",
                     nboots = 500, parallel = TRUE, cores = 16,
                     placeboTest = TRUE,
                     placebo.period = c(-2, 0))
```

```{r placebo-ife-plot}
plot(out_ife_pb, main = "Placebo test --- IFE-EM on simdata")
```

The numeric output for the placebo region is in `out_ife_pb$est.placebo`.

A note on what `gsynth()` does *not* expose: **fect** also has a `carryoverTest` argument that holds out late post-treatment periods to check whether treatment effects persist past the end of the treatment window. That test requires panels with treatment reversals, which the GSC framework does not allow, so `gsynth()` does not accept `carryoverTest`. If you have a panel with reversals and want a carryover test, use `fect::fect()` directly.

## Matrix completion on unbalanced data

Applying matrix completion to the unbalanced `turnout` data:

```{r}
set.seed(123456)
turnout.ub <- turnout[-c(which(turnout$abb == "WY")[1:15],
                         sample(1:nrow(turnout), 50, replace = FALSE)), ]
```

```{r turnout-mc, cache = TRUE}
out_mc_ub <- gsynth(turnout ~ policy_edr + policy_mail_in + policy_motor,
                    min.T0 = 8, data = turnout.ub,
                    index = c("abb", "year"), estimator = "mc",
                    se = TRUE, inference = "bootstrap",
                    nboots = 1000, seed = 02139,
                    parallel = TRUE, cores = 16)
```

```{r turnout-mc-plot}
plot(out_mc_ub, main = "Estimated ATT (MC, unbalanced turnout)",
     xlim = c(-10, 5), ylim = c(-10, 20))
```

## Migration to fect {#sec-migration}

When v2.0.0 ships (target: late 2026), the IFE-EM and MC paths in `gsynth` will be removed entirely. Migrating now means moving these calls to `fect::fect()`. The substitution is mechanical.

### IFE-EM migration

```{r ife-migration, eval = FALSE}
## v1.4.0:
gsynth(Y ~ D + X1 + X2, data = simdata,
       index = c("id", "time"), estimator = "ife",
       inference = "bootstrap", se = TRUE,
       nboots = 200, r = c(0, 5), CV = TRUE,
       force = "two-way")

## v1.5.0+ (use fect directly):
fect::fect(Y ~ D + X1 + X2, data = simdata,
           index = c("id", "time"),
           method = "ife",
           time.component.from = "notyettreated",
           vartype = "bootstrap", se = TRUE,
           nboots = 200, r = c(0, 5), CV = TRUE,
           force = "two-way")
```

### Matrix completion migration

```{r mc-migration, eval = FALSE}
## v1.4.0:
gsynth(Y ~ D + X1 + X2, data = simdata,
       index = c("id", "time"), estimator = "mc",
       inference = "bootstrap", se = TRUE, nboots = 200,
       force = "two-way")

## v1.5.0+ (use fect directly):
fect::fect(Y ~ D + X1 + X2, data = simdata,
           index = c("id", "time"),
           method = "mc",
           time.component.from = "notyettreated",
           vartype = "bootstrap", se = TRUE, nboots = 200,
           force = "two-way")
```

### `EM = TRUE` migration

`EM = TRUE` was an early flag that mapped to IFE-EM. In v1.5.0, the canonical replacement is `fect::fect(method = "ife", ...)` directly:

```{r em-migration, eval = FALSE}
## v1.4.0:
gsynth(Y ~ D + X1 + X2, data = simdata,
       index = c("id", "time"), EM = TRUE,
       se = TRUE, nboots = 200, r = c(0, 5))

## v1.5.0+:
fect::fect(Y ~ D + X1 + X2, data = simdata,
           index = c("id", "time"),
           method = "ife",
           time.component.from = "notyettreated",
           vartype = "bootstrap", se = TRUE,
           nboots = 200, r = c(0, 5))
```

### Argument mapping

| `gsynth()` arg | `fect::fect()` equivalent |
| --- | --- |
| `inference = "parametric"` | `vartype = "parametric"` |
| `inference = "nonparametric"` | `vartype = "bootstrap"` |
| `inference = "bootstrap"` | `vartype = "bootstrap"` |
| `inference = "jackknife"` | `vartype = "jackknife"` |
| `weight = "col"` | `W = "col"` |
| `EM = TRUE` | `method = "ife"`, `time.component.from = "notyettreated"` |
| `estimator = "ife"` | `method = "ife"`, `time.component.from = "notyettreated"` |
| `estimator = "mc"` | `method = "mc"`, `time.component.from = "notyettreated"` |
| `estimator = "gsynth"` | `method = "ife"` (or `"gsynth"`), `time.component.from = "nevertreated"` |

Most other arguments (`CV`, `r`, `force`, `nboots`, `parallel`, `cores`, `seed`, `min.T0`, `alpha`, `normalize`, `tol`) carry over by name.

If you maintain a replication archive that depends on `gsynth(..., estimator = "ife")` or `gsynth(..., estimator = "mc")`, plan for one of two paths:

1. **Pin to gsynth 1.4.0** via the CRAN archive: `remotes::install_version("gsynth", "1.4.0")`.
2. **Migrate the calls to `fect::fect()`** using the recipes above.

After v2.0.0, bug fixes and feature work for IFE-EM and MC happen exclusively in **fect**.
