# Welcome {.unnumbered}

This manual serves as a user guide for the **gsynth** package in R.

Starting from v1.3.0, **gsynth** becomes a wrapper of **fect**. We recommend users consult the **fect** [User Manual](https://yiqingxu.org/packages/fect/) for the most up-to-date features. We maintain **gsynth** for backward compatibility. For the technical details of the GSC estimator, see @Xu2017.

::: {.callout-tip}

## What's new in v1.5.0

- **fect 2.4.x catch-up.** 
  - new rolling-window cross-validation defaults (`cv.method = "rolling"`, `k = 20`); 
  - new `ci.method` argument for the confidence intervals reported in `est.*` slots; 
  - new `placeboTest` and `placebo.period` arguments for placebo testing; 
  - new `legacy.style`, `highlight`, and `highlight.fill` arguments on `plot.gsynth()`;
  - loading-overlap plot type available via `plot(fit, type = "loading.overlap")`.
- **Soft-deprecated paths.** 
  - `estimator = "ife"` (IFE-EM), `estimator = "mc"` (matrix completion), `EM = TRUE`, and `effect()`. Each emits a one-time-per-session deprecation message. 
  - For IFE-EM, matrix completion, and post-hoc estimands (cumulative ATT, APTT, log-ATT), use [`fect::fect()`](https://yiqingxu.org/packages/fect/) and `fect::estimand()` directly. 
- **Breaking change.** `inference = "parametric"` with `estimator = "ife"` or `"mc"` now errors instead of silently coercing to bootstrap. See [Chapter @sec-ife-mc].

:::

## Installation

**gsynth** v1.5.0 requires **fect** `>= 2.4.2`. Until **fect** v2.4.2 is on CRAN, install both the **fect** development branch and **gsynth** from GitHub:

```{r install-fect-dev, eval = FALSE}
# Up-to-date fect from the dev branch (>= 2.4.2)
devtools::install_github("xuyiqing/fect", ref = "dev")
```

```{r install-github, eval = FALSE, message = FALSE, warning = FALSE}
# gsynth development version
devtools::install_github("xuyiqing/gsynth")
```

Once **fect** v2.4.2 is on CRAN, the standard install will suffice:

```{r install-cran, eval = FALSE}
install.packages("gsynth")
```

```{r check-version}
# Check version
installed.packages()["gsynth", "Version"]
```

```{r load}
# Load package
library(gsynth)
```

## Quick reference

**gsynth** has one main function, `gsynth()`, which estimates causal effects using interactive fixed-effect models. Three estimation methods are nominally available, but only the GSC path is recommended in v1.5.0+:

| `estimator =` | Status | Method | Reference |
| --- | --- | --- | --- |
| `"gsynth"` (default) | **recommended** | Generalized Synthetic Control | @Xu2017 |
| `"ife"` | soft-deprecated; use `fect::fect(method = "ife", ...)` | IFE-EM | @Gobillon2016 |
| `"mc"` | soft-deprecated; use `fect::fect(method = "mc", ...)` | Matrix completion | @Athey2021 |

The legacy `EM = TRUE` parameter is also soft-deprecated and equivalent to `estimator = "ife"`.

### Key arguments

| Argument | Type | Description |
| --- | --- | --- |
| `formula` | formula | `Y ~ D + X1 + X2`; first RHS variable is the treatment |
| `data` | data.frame | Long-format panel data |
| `index` | character(2) | `c("unit_var", "time_var")` |
| `estimator` | character | `"gsynth"` (recommended); `"ife"` / `"mc"` soft-deprecated |
| `force` | character | `"none"` / `"unit"` / `"time"` / `"two-way"` |
| `r` | integer(2) | Range for number of factors, e.g. `c(0, 5)` |
| `CV` | logical | Cross-validation to select number of factors |
| `cv.method` | character | `"rolling"` (default) or `"block"` |
| `cv.prop` | numeric | Per-fold unit-sampling fraction (default 0.1) |
| `cv.buffer` | integer | Past-side buffer for rolling CV (default 1) |
| `k` | integer | CV folds (default 20 in v1.5.0+) |
| `time.component.from` | character | `"nevertreated"` (GSC) or `"notyettreated"` (IFE-EM/MC) |
| `se` | logical | Compute uncertainty estimates |
| `inference` | character | `"parametric"` / `"nonparametric"` / `"jackknife"` |
| `ci.method` | character | `"normal"` (default; Wald) or `"basic"` (reflected pivot). New in v1.5.0. |
| `nboots` | integer | Number of bootstrap runs (default 200; bump to 1000+ for tail-quantile CIs) |
| `placeboTest` | logical | Hold out pre-treatment periods as placebo. New in v1.5.0. |
| `placebo.period` | integer | Event-time periods used for the placebo block, e.g. `c(-2, 0)` |
| `weight`, `W.est`, `W.agg` | character | Weight column(s); per-role variants for outcome fit / aggregation |
| `parallel` | logical | Enable parallel computing |
| `cores` | integer | Number of cores (default: auto-detect) |
| `min.T0` | integer | Minimum pre-treatment periods for treated units |

## Organization

- [Chapter @sec-gsynth] --- Generalized Synthetic Control: block DID, inference, placebo test, loading-overlap diagnostic, staggered DID, unbalanced panels
- [Chapter @sec-ife-mc] --- IFE-EM and matrix completion (soft-deprecated); placebo test for these methods; migration to `fect::fect()`
- [Chapter @sec-plots] --- Plot customization, modern theme, highlight API, plot types (counterfactual, factors, loadings, loading-overlap, raw, missing)

## Contributors

- [Yiqing Xu](https://yiqingxu.org/){target="_blank"}
- [Licheng Liu](https://liulch.github.io/){target="_blank"}
- Ziyi Liu
- Shiyun Hu

## How to Cite

To cite the **gsynth** package, please use:

> Xu, Yiqing. 2017. "Generalized Synthetic Control Method: Causal Inference with Interactive Fixed Effects Models." *Political Analysis* 25 (1): 57--76.

```bibtex
@article{Xu2017,
  title = {Generalized Synthetic Control Method: Causal Inference with Interactive Fixed Effects Models},
  author = {Xu, Yiqing},
  journal = {Political Analysis},
  volume = {25},
  number = {1},
  pages = {57--76},
  year = {2017}
}
```

For technical background on the v2.0–v2.4 changes in **fect** (which `gsynth` v1.5.0 inherits), see the **fect** [User Manual](https://yiqingxu.org/packages/fect/).

## Report Bugs

Please report bugs to yiqingxu [at] stanford.edu or open an issue on [GitHub](https://github.com/xuyiqing/gsynth/issues).

<!-- badges: start -->

[![Lifecycle: stable](https://lifecycle.r-lib.org/articles/figures/lifecycle-stable.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable) [![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT) [<img src="https://www.r-pkg.org/badges/version/gsynth" alt="CRAN status"/>](https://CRAN.R-project.org/package=gsynth) [<img src="https://cranlogs.r-pkg.org/badges/grand-total/gsynth" alt="downloads: CRAN"/>](https://cran.r-project.org/package=gsynth)

<!-- badges: end -->
