# Welcome {.unnumbered}

This manual serves as a user guide for the **gsynth** package in R. 

Starting from v1.3.0, **gsynth** becomes a wrapper of **fect**. We recommend users consult the **fect** [User Manual](https://yiqingxu.org/packages/fect/) for the most up-to-date features. We maintain **gsynth** for backward compatibility. For the technical details, see @Xu2017.

## Installation

```{r install-cran, eval = FALSE}
# From CRAN
install.packages("gsynth")
```

```{r install-github, eval = FALSE, message = FALSE, warning = FALSE}
# Development version
devtools::install_github("xuyiqing/gsynth")
```

```{r check-version}
# Check version
installed.packages()["gsynth", "Version"]
```

```{r load}
# Load package
library(gsynth)
```

## Quick-reference

**gsynth** has one main function, `gsynth()`, which estimates causal effects using interactive fixed-effect models. Three estimation methods are available:

| `estimator =` | Method | Algorithm | Reference |
| --- | --- | --- | --- |
| `"gsynth"` (default) | Generalized Synthetic Control | Factors estimated from control group only | @Xu2017 |
| `"ife"` | Interactive Fixed Effects with EM | EM algorithm using treated units' pre-treatment data | @Gobillon2016 |
| `"mc"` | Matrix Completion | Nuclear norm regularization | @Athey2021 |

The legacy `EM = TRUE` parameter is equivalent to `estimator = "ife"`.

### Key arguments

| Argument | Type | Description |
| --- | --- | --- |
| `formula` | formula | `Y ~ D + X1 + X2`; first RHS variable is the treatment |
| `data` | data.frame | Long-format panel data |
| `index` | character(2) | `c("unit_var", "time_var")` |
| `estimator` | character | `"gsynth"` / `"ife"` / `"mc"` |
| `force` | character | `"none"` / `"unit"` / `"time"` / `"two-way"` |
| `r` | integer(2) | Range for number of factors, e.g. `c(0, 5)` |
| `CV` | logical | Cross-validation to select number of factors |
| `se` | logical | Compute uncertainty estimates |
| `inference` | character | `"parametric"` / `"nonparametric"` / `"jackknife"` |
| `nboots` | integer | Number of bootstrap runs |
| `parallel` | logical | Enable parallel computing |
| `cores` | integer | Number of cores (default: auto-detect) |
| `min.T0` | integer | Minimum pre-treatment periods for treated units |



## Organization

- [Chapter @sec-gsynth] — Generalized Synthetic Control (block DID, staggered DID, unbalanced panels)
- [Chapter @sec-ife-mc] — IFE-EM and Matrix Completion methods

## Contributors

- [Yiqing Xu](https://yiqingxu.org/){target="_blank"}
- [Licheng Liu](https://liulch.github.io/){target="_blank"}
- Ziyi Liu
- Shiyun Hu

## Report bugs

Please report bugs to yiqingxu [at] stanford.edu or open an issue on [GitHub](https://github.com/xuyiqing/gsynth/issues).

<!-- badges: start -->

[![Lifecycle: stable](https://lifecycle.r-lib.org/articles/figures/lifecycle-stable.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable) [![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT) [<img src="https://www.r-pkg.org/badges/version/gsynth" alt="CRAN status"/>](https://CRAN.R-project.org/package=gsynth) [<img src="https://cranlogs.r-pkg.org/badges/grand-total/gsynth" alt="downloads: CRAN"/>](https://cran.r-project.org/package=gsynth)

<!-- badges: end -->
