---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gsynth

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-green.svg)](https://www.tidyverse.org/lifecycle/#stablel)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

**gsynth** implements the [generalized synthetic control method](https://doi-org.stanford.idm.oclc.org/10.1017/pan.2016.2), which imputes counterfactuals for each treated unit using control group information based on a linear interactive fixed effects model.

**Authors:** [Yiqing Xu](https://yiqingxu.org/) (Stanford), [Licheng Liu](https://liulch.github.io/) (UMich), Ziyi Liu (Berkeley), Shiyun Hu (PKU)

**Date:** Dec 7, 2025

**Repos:** [Github](https://github.com/xuyiqing/gsynth) (1.3.1)
[CRAN](https://cran.r-project.org/web/packages/gsynth/index.html) (1.3.1)

**Example:** R code used in the [tutorial](https://yiqingxu.org/packages/gsynth/articles/tutorial.html) can be downloaded from [here](https://yiqingxu.org/packages/gsynth/gsynth_examples.R).

---

## Installation

You can install **gsynth** directly from CRAN by typing the following command in the **R** console: 

```{r eval=FALSE}
install.packages('gsynth')
```

You can also install the development version of the package from Github by typing:
```{r eval=FALSE}
install.packages('devtools', repos = 'http://cran.us.r-project.org') # if not already installed
devtools::install_github('xuyiqing/gsynth')
```

**gsynth** depends on the following packages, which will be installed automatically when **gsynth** is being installed; you can also install them manually:  
```{r eval=FALSE}
packages <- c("ggplot2", "panelViews", "fect")

# install packages if not already installed
install_all <- function(packages) {
  installed_pkgs <- installed.packages()[, "Package"]
  for (pkg in packages) {
    if (!pkg %in% installed_pkgs) {
      install.packages(pkg)
    }
  }
}
install_all(packages)
```

## Report bugs

**gsynth** is now a wrapper of **fect**. Please check [fect User Manual](https://yiqingxu.org/packages/fect/) for updates. We maintain **gsynth** solely for backward compatibility.


